---
title: "COVID-19"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
---

*Purpose*: We can't *possibly* do a class on data science and *not* look at covid-19. Come on.

In this challenge, you'll learn how to navigate the U.S. Census Bureau website, programmatically download data from the internet, and perform a county-level population-weighted analysis of current covid-19 trends. Get excited!

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)

```

*Background*: [COVID-19](https://en.wikipedia.org/wiki/Coronavirus_disease_2019) is the disease caused by the virus SARS-CoV-2. In 2020 it became a global pandemic, leading to huge loss of life and tremendous disruption to society. The New York Times published up-to-date data on the progression of the pandemic across the United States---we will study these data in this challenge.

# The Big Picture
<!-- -------------------------------------------------- -->

We're about to go through *a lot* of weird steps, so let's first fix the big picture firmly in mind:

We want to study COVID-19 in terms of data: both case counts (number of infections) and deaths. We're going to do a county-level analysis in order to get a high-resolution view of the pandemic. Since US counties can vary widely in terms of their population, we'll need population estimates in order to compute infection rates (think back to the `Titanic` challenge).

That's the high-level view; now let's dig into the details.

# Get the Data
<!-- -------------------------------------------------- -->

1. County-level population estimates (Census Bureau)
2. County-level COVID-19 counts (New York Times)

## Get Census Data of Population


```{r import census data}
## TASK: Load the census bureau data with the following tibble name.
df_pop  <-  read_csv("data/ACSST5Y2018.S0101_data_with_overlays_2020-08-01T094026.csv",skip = 1)  

df_pop_select <- df_pop %>% 
  select(id,Geographic_Area_Name, EstimatePopulation) %>% 
  separate(
    col = 2,
    into = c("county", NA, "state"),
    sep = " "
  )

#df_pop_select
```

*Note*: You can find information on 1-year, 3-year, and 5-year estimates [here](https://www.census.gov/programs-surveys/acs/guidance/estimates.html). The punchline is that 5-year estimates are more reliable but less current.

## Automated Download of NYT Data
<!-- ------------------------- -->




Once you have the url, the following code will download a local copy of the data, then load the data into R.

```{r nyt data download}
url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
n
## NOTE: No need to change this; just execute
## Set the filename of the data to download
filename_nyt <- "./data/nyt_counties.csv"

## Download the data locally
curl::curl_download(
        url_counties,
        destfile = filename_nyt
      )

## Loads the downloaded csv
df_covid <- read_csv(filename_nyt)
```

You can now re-run the chunk above (or the entire notebook) to pull the most recent version of the data. Thus you can periodically re-run this notebook to check in on the pandemic as it evolves.

*Note*: You should feel free to copy-paste the code above for your own future projects!

# Join the Data
<!-- -------------------------------------------------- -->



To get a sense of our task, let's take a glimpse at our two data sources.

```{r glimpse}  
## NOTE: No need to change this; just execute
df_pop_select %>% glimpse
df_covid %>% glimpse
```

To join these datasets, we'll need to use [FIPS county codes](https://en.wikipedia.org/wiki/FIPS_county_code).[2] The last `5` digits of the `id` column in `df_pop` is the FIPS county code, while the NYT data `df_covid` already contains the `fips`.

__q3__ Process the `id` column of `df_pop` to create a `fips` column.

```{r q3-task}
## TASK: Create a `fips` column by extracting the county code
df_q3 <- df_pop_select %>% 
  mutate(fips = substr(id,10,15))

df_q3

  
  
```

```{r fips test}
df_q3 %>%
  filter(county == "Autauga") %>%
  pull(fips) == "01001"



```

Use the following test to check your answer.

```{r q3-tests}
## NOTE: No need to change this
## Check known county
assertthat::assert_that(
              (df_q3 %>%
              filter(county == "Autauga") %>%
              pull(fips)) == "01001"
              )
print("Very good!")
```

```{r}
df_q3
```


__q4__ Join `df_covid` with `df_q3` by the `fips` column. Use the proper type of join to preserve all rows in `df_covid`.

```{r q4-task}
## TASK: Join df_covid and df_q3 by fips.
df_q4 <- merge(df_covid, df_q3, by = c("fips","county","state")) #all.y = TRUE)

#df_q4[is.na(df_q4)] <- 0



```


For convenience, I down-select some columns and produce more convenient column
names.

```{r rename}
## NOTE: No need to change; run this to produce a more convenient tibble
df_data <-
  df_q4 %>%
  select(
    date,
    county,
    state,
    fips,
    cases,
    deaths,
    population = `EstimatePopulation`
  )
```


# Analyze
<!-- -------------------------------------------------- -->

Now that we've done the hard work of loading and wrangling the data, we can finally start our analysis. Our first step will be to produce county population-normalized cases and death counts. Then we will explore the data.

## Normalize
<!-- ------------------------- -->

__q5__ Use the `population` estimates in `df_data` to normalize `cases` and `deaths` to produce per 100,000 counts.[3] Store these values in the columns `cases_perk` and `deaths_perk`.

```{r q5-task}
## TASK: Normalize cases and deaths
df_normalized <-
  df_data %>% 
  mutate(cases_perk = cases/(population/100000), deaths_perk = deaths/(population/100000)) 

```

```{r q5-tests}
## NOTE: No need to change this
## Check known county data
assertthat::assert_that(
              abs(df_normalized %>%
               filter(str_detect(county, "Snohomish"),
                 date == "2020-01-21"
               ) %>%
              pull(cases_perk) - 0.127) < 1e-3
            )
assertthat::assert_that(
              abs(df_normalized %>%
               filter(
                 str_detect(county, "Snohomish"),
                 date == "2020-01-21"
               ) %>%
              pull(deaths_perk) - 0) < 1e-3
            )

print("Excellent!")
```

## Guided EDA
<!-- ------------------------- -->

Before turning you loose, let's complete a couple guided EDA tasks.

__q6__ Compute the mean and standard deviation for `cases_perk` and `deaths_perk`.

```{r q6-task}
df_normalized %>% 
  summarise(mean_cases = mean(cases_perk), sd_cases = sd(cases_perk), mean_deaths = mean(deaths_perk), sd_deaths = sd(deaths_perk))

```

__q7__ Find the top 10 counties in terms of `cases_perk`, and the top 10 in terms of `deaths_perk`. Report the population of each county along with the per-100,000 counts. Compare the counts against the mean values you found in q6. Note any observations. Does New York City show up in the top? Why or why not?

```{r q7-task}
## TASK: Find the top 10 max cases_perk counties; report populations as well
df_normalized %>% 
  group_by(county,state) %>% 
  summarise(max_cases_perk = max(cases_perk), population = max(population)) %>% 
  ungroup() %>% 
  arrange(desc(max_cases_perk))# %>% 
  #head(10)

#df_data %>% 
#  filter(county == "Fairfax") %>% 
#  arrange(date)

## TASK: Find the top 10 deaths_perk counties; report populations as well
```

**Observations**:

New York is not on this list. It looks like most of these locations are smaller towns that have had outbreaks in institutions where physical distancing is difficult. For example:

Trousdale - outbreak in a prison https://communityimpact.com/nashville/southwest-nashville/coronavirus/2020/05/01/large-spike-in-coronavirus-cases-traced-to-prison-in-trousdale-county/

Lake Tennessee: outbreak at a prison https://www.wate.com/health/coronavirus/tennessees-lake-county-leads-us-in-per-capita-virus-cases/

Lee Arkansas: outbreak at a prison
https://wreg.com/news/small-arkansas-county-dealing-with-rise-in-covid-19-cases/

Dakota Nebraska: outbreak at a meatpacking plant: 
https://omaha.com/news/state_and_regional/786-workers-at-tysons-dakota-city-plant-have-coronavirus-company-says-worst-is-over/article_97279b04-9376-5990-861e-c423c47f13ab.html

Lincoln Arkansas: Outbreak at a prison.

Nobles Minnesota - outbreak at a pork plant

Colfax Nebraska - outbreak at a meatpacking plant

Chattahoochee Georgia - outbreak at Fort Bennings
https://www.georgiahealthnews.com/2020/06/fort-benning-cases-covid-19-surge-chattahoochee-county/

Ford Kansas - uncertain 

Lowndes Alabama - uncertain 

## Self-directed EDA
<!-- ------------------------- -->

__q8__ Drive your own ship: You've just put together a very rich dataset; you now get to explore! Pick your own direction and generate at least one punchline figure to document an interesting finding. 

Multiple outbreaks occurred in prisons. Are prisons driving case loads in other parts of the US?

## Investigating Institutional Data

Import data from 2010 US Census to understand state-wise institutionalized populations.

```{r}
prison_pop = read_csv("data/prison_pop2.csv",skip = 2)

prison_pop2 = prison_pop %>% 
  mutate(fips = substr(id,10,15)) %>% 
  select(
    fips,
    Geographic_Area_Name,
    total = Total,
    total_institutional,
    total_correctional,
    nursing_home,
    college_housing,
    miliary_housing
   ) %>% 
  separate(
    col = 2,
    into = c("county", NA, "state"),
    sep = " "
  ) 

```


```{r institution normalize}
institutional_covid <- merge(df_data, prison_pop2, by = c("fips","county","state"))

institutional_normalized <- institutional_covid %>% 
  mutate(popk = population/100000) %>% 
  mutate(cases_perk = cases/popk, deaths_perk = deaths/popk) %>% 
  mutate(total_perk = total/popk, total_institutional_perk = total_institutional / popk, total_correctional_perk = total_correctional / popk) %>% 
  mutate(nursing_home_perk = nursing_home / popk, college_perk = college_housing / popk, military_perk = miliary_housing / popk)

max_cases <- institutional_normalized %>% 
  group_by(county,state,fips) %>% 
  summarise(
    max_cases = max(cases),
    max_cases_perk = max(cases_perk), 
    max_deaths = max(deaths),
    max_deaths_perk = max(deaths_perk), 
    population = max(population), total = max(total), 
    total = max(total),
    total_perk = max(total_perk),
    total_institutional = max(total_institutional),
    total_institutional_perk = max(total_institutional_perk),
    total_correctional = max(total_correctional),
    total_correctional_perk = max(total_correctional_perk),
    nursing_home = max(nursing_home),
    nursing_home_perk = max(nursing_home_perk),
    college_housing = max(college_housing),
    college_perk = max(college_perk),
    military_housing = max(miliary_housing),
    military_perk = max(military_perk)
  ) %>%
  ungroup() %>% 
  arrange(desc(max_cases_perk))

```

Now, let's look the number of cases vs. the institutionalized population per county (with both quantities normalized by population).


```{r bubble-institutionalized}
library(ggrepel)

label_size = 2

outlier_cases <- 
  max_cases  %>%
    filter(max_cases_perk >= 7500) %>%
    unite("county_state", county:state, sep=", ", remove = FALSE)

outlier_institutional <-
  max_cases  %>%
    filter(total_institutional_perk > 30000) %>%
    unite("county_state", county:state, sep=", ", remove = FALSE)

max_cases %>% 
  unite("county_state", county:state, sep=", ", remove = FALSE) %>%
  arrange(desc(population)) %>% 
  ggplot() +
  geom_point(aes(x = total_institutional_perk, y = max_cases_perk, size = max_deaths_perk),alpha = 0.2, color = "blue") + 
  geom_label_repel(size = label_size, data = outlier_cases, aes(x = total_institutional_perk, y = max_cases_perk, label = county_state, fill = "max_cases_perk > 7500")) +
  geom_label_repel(size = label_size, data = outlier_institutional, aes(x = total_institutional_perk, y = max_cases_perk, label = county_state, fill = "total_institutional_perk > 30000")) +
  #geom_smooth(aes(x = nursing_home, y = max_deaths)) +
  scale_size(range = c(.1, 8), name="Deaths Per 100k") 
  # coord_cartesian(xlim = c(0,1500))
```

Observations:

- Greensville, Virginia - among the cyan-labeled institutionalized counties, this has the largest Deaths per 100k.
    - Home to the [Greensville Correctional Center](https://en.wikipedia.org/wiki/Greensville_Correctional_Center), which had a [small outbreak in May](https://www.wtvr.com/news/coronavirus/family-members-concerned-after-246-cases-of-covid-19-at-greensville-correctional-center)
    - Does this indicate something peculiar about VA?
- Counties in Tennessee are outliers with respect to case counts (`max_cases_perk`).

## Virginia Institutional Case Counts
    
Let's visualize the county-level case counts while highlighting Greensville. 
  
```{r plot-virginia}
library(urbnmapr)
library(sf)

target_date = "2020-08-05"

df_normalized_county <- df_normalized
names(df_normalized_county)[names(df_normalized_county) == "fips"] <- "county_fips"

isolated_county_fips <-
  df_normalized_county %>%
  filter(state == "Virginia", county == "Greensville") %>%
  summarize(fips = unique(county_fips))

isolated_county <-
  urbnmapr::counties %>%
  filter(county_fips == isolated_county_fips$fips)

isolated_state <-
  urbnmapr::states %>%
  filter(state_name == "Virginia")

df_normalized_county %>%
  filter(state == "Virginia", date == target_date) %>%
  left_join(counties, by = "county_fips") %>%
  ggplot() +
  geom_polygon(aes(long, lat, group = group), data = isolated_state, fill = "grey", color = "white") +
  geom_polygon(aes(long, lat, group = group, fill = cases_perk), color = "#ffffff", size = .25) +
  geom_polygon(aes(long, lat, group = group), data = isolated_county, fill = NA, color = "red") +
  coord_map(projection = "albers", lat0 = 36, lat1 = 42)


```

Greensville (outlined in red) stands out among counties in VA. Does the county have similarly high death counts?

```{r plot-virginia-deaths}
library(urbnmapr)
library(sf)

target_date = "2020-08-05"

isolated_county_fips <-
  df_normalized_county %>%
  filter(state == "Virginia", county == "Greensville") %>%
  summarize(fips = unique(county_fips))

isolated_county <-
  urbnmapr::counties %>%
  filter(county_fips == isolated_county_fips$fips)

isolated_state <-
  urbnmapr::states %>%
  filter(state_name == "Virginia")

df_normalized_county <- df_normalized
names(df_normalized_county)[names(df_normalized_county) == "fips"] <- "county_fips"

df_normalized_county %>%
  filter(state == "Virginia", date == target_date) %>%
  left_join(counties, by = "county_fips") %>%
  ggplot() +
  geom_polygon(aes(long, lat, group = group), data = isolated_state, fill = "grey", color = "white") +
  geom_polygon(aes(long, lat, group = group, fill = deaths_perk), color = "#ffffff", size = .25) +
  geom_polygon(aes(long, lat, group = group), data = isolated_county, fill = NA, color = "red") +
  coord_map(projection = "albers", lat0 = 36, lat1 = 42)

```

The normalized death counts in Greensville are lower than other VA counties despite higher case counts. Could this mean institutionalized populations are receiving high quality (or at least decent) medical treatment?

## Tennessee Institutional Case Counts

Let's look at similar plots for Tennessee. Here, we highlight counties with 2000 or more institutionalized persons perk 10k.

```{r tennessee-map}
library(urbnmapr)
library(sf)
library(maps)

df_normalized_county <- df_normalized
names(df_normalized_county)[names(df_normalized_county) == "fips"] <- "county_fips"

max_cases_county <- max_cases
names(max_cases_county)[names(max_cases_county) == "fips"] <- "county_fips"

isolated_county_fips <-
  max_cases_county %>%
  mutate(prison = total_correctional_perk > 2000 | county == "Trousdale County") %>% 
  filter(state == "Tennessee", prison == TRUE) %>%
  summarize(fips = unique(county_fips))

isolated_county <-
  urbnmapr::counties %>%
  filter(county_fips == isolated_county_fips$fips)

max_cases_county %>%
  filter(state == "Tennessee") %>%
  #filter(date == "2020-08-05") %>%
  left_join(counties, by = "county_fips") %>%
  ggplot(aes(long, lat, group = group, fill = max_cases_perk)) +
  geom_polygon(color = "#ffffff", size = .1) +
  coord_map(projection = "albers", lat0 = 36, lat1 = 42) + 
  geom_polygon(aes(long, lat, group = group), data = isolated_county, fill = NA, color = "red")
```

Trousdale county is not outlined highlighted since the prison was not opened until after 2010 and the above plots rely on 2010 Census data. Regardless, it clearly stands out as the largest outbreak in Tennessee.


## Summary of COVID Cases in Prisons per State

The Prison Policy Initiative publishes data on case counts, death counts, test numbers, and prison populations at a state level. Here, we look at their data.

```{r prison-cases}

# source: https://www.prisonpolicy.org/blog/2020/06/24/covidrates/
library(readr)
df_prison_covid <- read_csv("data/prison_policy_covid.csv")

glimpse(df_prison_covid)
```

The case counts (below) show that New Jersey, Tennessee, and Michigan have the largest case counts among institutionalized populations per 10k.

```{r policy-plot-cases}
df_prison_covid %>%
  # mutate_each_(funs(as.numeric), Deaths_per_10k)
  filter(cases_per_10k > 0) %>%
  arrange(desc(cases_per_10k)) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(State, -cases_per_10k), cases_per_10k), stat = "identity") +
  theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 8)) +
  labs(x = "State")
```

The death counts (below) show that New Jersey, Michigan, and Ohio have the largest death counts among institutionalized populations per 10k.

```{r policy-plot-deaths}
df_prison_covid %>%
  # mutate_each_(funs(as.numeric), Deaths_per_10k)
  filter(Deaths_per_10k > 0) %>%
  arrange(desc(Deaths_per_10k)) %>%
  ggplot() +
  geom_bar(mapping = aes(x = reorder(State, -Deaths_per_10k), Deaths_per_10k), stat = "identity") +
  theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 8)) +
  labs(x = "State")
```

# ```{r policy-case-fatality-rate}
# df_prison_covid %>%
#   filter(Deaths_per_confirmed_case > 0) %>%
#   mutate(case_fatality_rate = as.numeric(sub("%", "", Deaths_per_confirmed_case))) %>%
#   arrange(desc(case_fatality_rate)) %>%
#   ggplot() +
#   geom_bar(mapping = aes(x = reorder(State, -case_fatality_rate), case_fatality_rate), stat = "identity") +
#   theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 7)) +
#   labs(x = "State", y = "Case Fatality Rate [%]")
# ```

How well are prisons reporting test results? The data from the Prison Policy Initiative shows a good deal of missingness, as grey states indicate no reported testing data from prisons.

```{r map-prison-tests}
df_prison_covid %>%
  mutate(
    N_tests = as.numeric(sub(",","",N_tests_given)),
    state_name = State
         ) %>%
  mutate(
   percent_pop_tested = 100 * N_tests / Prison_population
  ) %>%
  filter(N_tests > 0) %>%
  left_join(urbnmapr::states, by = "state_name") %>%
  glimpse() %>%
  ggplot() +
  geom_polygon(aes(long, lat, group = group), data = urbnmapr::states, fill = "grey", color = "white") +
  geom_polygon(aes(long, lat, group = group, fill = percent_pop_tested), color = "#ffffff", size = .25)
  # geom_polygon(aes(long, lat, group = group), data = isolated_county, fill = NA, color = "red") +
  coord_map(projection = "albers", lat0 = 36, lat1 = 42)
```

Prisons (and other institutions where physical distancing is difficult) should do better at A) testing and B) reporting test results to external stakeholders.


### Aside: Some visualization tricks
<!-- ------------------------- -->

These data get a little busy, so it's helpful to know a few `ggplot` tricks to help with the visualization. Here's an example focused on Massachusetts.

```{r ma-example}
## NOTE: No need to change this; just an example
df_normalized %>%
  filter(state == "Massachusetts") %>%

  ggplot(
    aes(date, cases_perk, color = fct_reorder2(county, date, cases_perk))
  ) +
  geom_line() +
  scale_y_log10(labels = scales::label_number_si()) +
  scale_color_discrete(name = "County") +
  theme_minimal() +
  labs(
    x = "Date",
    y = "Cases (per 100,000 persons)"
  )
```

*Tricks*:

- I use `fct_reorder2` to *re-order* the color labels such that the color in the legend on the right is ordered the same as the vertical order of rightmost points on the curves. This makes it easier to reference the legend.
- I manually set the `name` of the color scale in order to avoid reporting the `fct_reorder2` call.
- I use `scales::label_number_si` to make the vertical labels more readable.
- I use `theme_minimal()` to clean up the theme a bit.
- I use `labs()` to give manual labels.

### Geographic exceptions
<!-- ------------------------- -->

The NYT repo documents some [geographic exceptions](https://github.com/nytimes/covid-19-data#geographic-exceptions); the data for New York, Kings, Queens, Bronx and Richmond counties are consolidated under "New York City" *without* a fips code. Thus the normalized counts in `df_normalized` are `NA`. To fix this, you would need to merge the population data from the New York City counties, and manually normalize the data.


# Notes
<!-- -------------------------------------------------- -->

[1] The census used to have many, many questions, but the ACS was created in 2010 to remove some questions and shorten the census. You can learn more in [this wonderful visual history](https://pudding.cool/2020/03/census-history/) of the census.

[2] FIPS stands for [Federal Information Processing Standards](https://en.wikipedia.org/wiki/Federal_Information_Processing_Standards); these are computer standards issued by NIST for things such as government data.

[3] Demographers often report statistics not in percentages (per 100 people), but rather in per 100,000 persons. This is [not always the case](https://stats.stackexchange.com/questions/12810/why-do-demographers-give-rates-per-100-000-people) though!
