---
title: "Aluminum Data"
author: "Mason del Rosario"
date: 2020-07-23
output:
  github_document:
    toc: true
---

*Purpose*: When designing structures such as bridges, boats, and planes, the design team needs data about *material properties*. Often when we engineers first learn about material properties through coursework, we talk about abstract ideas and look up values in tables without ever looking at the data that gave rise to published properties. In this challenge you'll study an aluminum alloy dataset: Studying these data will give you a better sense of the challenges underlying published material values.

In this challenge, you will load a real dataset, wrangle it into tidy form, and perform EDA to learn more about the data.

<!-- include-rubric -->
```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a number of Aluminum plates to determine their [elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and [Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio). These are key quantities used in the design of structural members, such as aircraft skin under [buckling loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested plats of various thicknesses, and at different angles with respect to the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking)) direction.

# Loading and Wrangle
<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data form many sources. The `read_csv()` function will help us load the data for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient for reporting in a table, but are not ideal for analysis.

__q1__ Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, mu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>%
    pivot_longer(
      names_to = c(".value", "angle"),
      names_transform = list(angle = as.integer),
      names_sep = "_",
      cols = c(-thick, -alloy)
    ) %>%
    filter(E != -1)

df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "mu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA
<!-- -------------------------------------------------- -->

## Initial checks
<!-- ------------------------- -->

__q2__ Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own question that you'd like to answer about the data.

First, some basic definitions:
- $E = $ Elastic's modulus; measures stiffness of solid material and describes relationship between stress and strain
- $\mu = $ Poisson's ratio; measures transverse strain in proportion to axial strain 

```{r q2-task}
# basic summary of all cols
summary(df_stang_long)

# unique values in categorical vars (including angle)
unique(df_stang_long$thick)
unique(df_stang_long$alloy)
unique(df_stang_long$angle)
```

**Observations**:

- Is there "one true value" for the material properties of Aluminum?
    - No; `E` and `mu` are not constant for each angle/thickness
- How many aluminum alloys were tested? How do you know?
    - One; see output of `unique(df_stang_long)`
- What angles were tested?
    - 0, 45, 90
- What thicknesses were tested?
  - 0.022, 0.032, 0.064, 0.081
- Do `E`, `mu` vary with thickness?
  

## Visualize
<!-- ------------------------- -->

__q3__ Create a visualization to investigate your question from q1 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q1 here
df_stang_long %>%
  ggplot(aes(x = thick, y = E)) +
  facet_wrap(~ angle) +
  geom_point(aes(color = angle)) +
  geom_smooth(aes(color = angle))

df_stang_long %>%
  ggplot(aes(x = thick, y = mu)) +
  facet_wrap(~ angle) +
  geom_point(aes(color = angle)) +
  geom_smooth(aes(color = angle))
```

**Observations**:

- Both properties decrease with increasing thickness, but not linearly.
  - We could use samples at more thicknesses to fill this in more.


__q4__ Consider the following statement:

"A material's property (or material property) is an intensive property of some material, i.e. a physical property that does not depend on the amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a tested plate. Does the following graph support or contradict the claim that "elasticity `E` is an intensive material property." Why or why not? Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(mu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

- Does this graph support or contradict the claim above?
    - (Weakly) contradicts. Based on the data, `E` and `mu` vary with thickness.
    - It stands to reason that these material properties are described by distributions, and the oft reported values are merely statistics summarizing the distribution (e.g., the population mean).
- Is this evidence conclusive?
    - This could be due to variations in the test apparatus/experimental procedure.

# References
<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural alloys for large strains" (1946) Journal of Research of the National Bureau of Standards, (pdf link)[https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26, (link)[https://en.wikipedia.org/wiki/List_of_materials_properties]
