---
title: "Antibiotics"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
---

*Purpose*: To create an effective visualization, we need to keep our *purpose* firmly in mind. There are many different ways to visualize data, and the only way we can judge efficacy is with respect to our purpose.

In this challenge you'll visualize the same data in two different ways, aimed at two different purposes.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

```{r setup}
library(tidyverse)
library(ggrepel)

```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization
<!-- -------------------------------------------------- -->

## Purpose: Compare Effectiveness
<!-- ------------------------- -->

__q1__ Create a visualization of `df_antibiotics` that helps you to compare the effectiveness of the three antibiotics across all the bacteria reported. Can you make any broad statements about antibiotic effectiveness?

```{r q1-task}
library(reshape2)
df_antibiotics.melt <- 
  melt(df_antibiotics, id = "bacteria", variable.name = "antibiotic", value.name = "MCI") %>%
  filter(antibiotic %in% c("penicillin",	"streptomycin",	"neomycin")) %>%
  transform(MCI = as.numeric(MCI)) 

df_antibiotics.melt %>%
  ggplot() +
  geom_point(aes(x = bacteria, y = log10(MCI), color = antibiotic)) +
  geom_hline(yintercept = -1) +
  theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 6))

df_antibiotics.melt %>%
  filter(MCI <= 0.1) %>%
  group_by(antibiotic) %>%
  summarise(n_usable = n())
  
```

**Observations**:

- *Note*: An antibiotic is effective if it falls below the black line (`MCI <= 0.1`).
- Neomycin appears to be usable for the most bacteria (9) of the three antibiotics.

## Purpose: Categorize Bacteria
<!-- ------------------------- -->

The *genus* of a living organism is a human categorization, based on various characteristics of the organism. Since these categories are based on numerous factors, we will tend to see clusters if we visualize data according to relevant variables. We can use these visuals to categorize observations, and to question whether given categories are reasonable!

__q2__ Create a visualization of `df_antibiotics` that helps you to categorize bacteria according to the variables in the data. Document your observations on how how clusters of bacteria in the variables do---or don't---align with their *genus* classification.

```{r q2-task}
# split bacteria col
df_genera <-
df_antibiotics %>%
  pivot_longer(
    names_to = "antibiotic",
    values_to = "MCI",
    cols = c("penicillin",	"streptomycin",	"neomycin")
  ) %>%
  separate(bacteria, c("genus", "species"), " ")
  
df_genera %>%
  ggplot(aes(x = genus, y = log10(MCI), color = antibiotic)) +
  geom_point() +
  geom_hline(yintercept = -1) +
  theme(axis.text.x = element_text(angle = 45,  hjust=1, size = 8)) +
  facet_grid(gram ~ .)

# summarize genus-wise MCI
df_genera %>%
  filter(MCI <= 0.1) %>%
  group_by(genus, gram) %>%
  summarise(n_usable = n())
```

**Observations**:

- *Effect of Genus*: Genera with impractically high MCI for all antibiotics:
    - Aerobacter, Escherichia, Klebsiella, Mycobacterium, Pseudomonas
- *Effect of Gram*: Among bacteria that are susceptible to the antibiotics, `gram = positive` bacteria fall under the `MCI` threshold more often.
 
# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." /American Scientist/ (2009) [link](https://www.americanscientist.org/article/thats-funny)
